---
title: "Staphylococcal occupancy within the human nasal microbiota - DATA ANALYSIS"
author: "Anna Ingham & Duncan Ng, Statens Serum Institut, Copenhagen, Denmark"
output: 
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.time(), '%d %B %Y')`"
---


# Load libraries
```{r setup, message=FALSE, warning=FALSE}
library(sessioninfo)
library(vegan)
library(ggplot2)
library(cowplot)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(ggrepel)
library(ggforce)
library(philentropy)
library(dplyr)
library(ape)
library(stringr)
library(rcompanion)
library(ggalluvial) 
library(tibble)
library(diagram)
library(markovchain)
library(ggplotify)
library(dendextend)
library(scales)
library(factoextra)
library(cluster)
library(tidyr)
library(kableExtra)
#library(apeglm)
library(corrplot)
library(viridis)
library(rcompanion)
library(vcdExtra)
library(ggbeeswarm)
library(treemap)
library(ggdendro)
```


# Import main dataset
The count table loaded here consists of samples that passed >5000 *tuf* reads that are classified. The data is summarized at species level. The accompanying metadata file is also loaded.  
```{r}
count_table = readRDS("filtered_data/combined_table_filter.rds")
count_metadata <- read.csv("data_input/count_metadata.csv", row.names = 1)
```

## Convert to proportions
The data is converted to proportions as a normalisation step.
```{r}
count_table = as.data.frame(apply(count_table, 2, function(x) x/sum(x)*100))
```

## Metadata summary
In total 1863 samples will be used for the analysis, including samples from Denmark (cross-sectional and longitudinal) and Germany (cross-sectional).  
```{r}
count_metadata %>% group_by(population, data_type) %>% tally() %>% kable() %>% kable_classic()
```

## Load data that has been excluded due to low read count.   
These data consist of samples that did not reach the 5000 read requirement.  
```{r}
count_table_removed = readRDS("filtered_data/combined_table_removed.rds")
count_meta_removed = readRDS("filtered_data/combined_meta_removed_with_bact_CST.rds")
count_meta_removed %>% group_by(population, data_type) %>% tally() %>% kable() %>% kable_classic()
```

# Distribution of species in cross-sectional data  
We focus on the cross-sectional data to get a summary of the staphylococcal species distribution.  
```{r}
frequency_meta = droplevels(count_metadata[which(count_metadata$data_type=="cross sectional"),])
frequency_table = count_table[,match(row.names(frequency_meta),colnames(count_table))]
species_summary = data.frame(
  row.names = row.names(frequency_table),
  sample_presence = apply(frequency_table, MARGIN = 1, function(x) length(which(x>0))),
  median = round(apply(frequency_table, MARGIN = 1, function(x) median(x)),digits = 2),
  min = round(apply(frequency_table, MARGIN = 1, function(x) min(x)),digits = 2),
  max = round(apply(frequency_table, MARGIN = 1, function(x) max(x)), digits = 2))
species_summary$percentage = round(species_summary$sample_presence/(length(colnames(frequency_table))) *100,digits = 2)
species_summary %>% kable() %>% kable_classic()
```


### Define color codes
```{r}
sCST_colors <- c("sCST1Sa_H" = "#e68502", "sCST1Sa_M" = "#f2ad00", "sCST2Sa" = "#fad77b", "sCST2Sc_H" = "#644b63", "sCST2Sl" = "#a7bf90" , "sCST2Sh" = "#385338", "sCST2Sc" = "#ac8cae", "sCST2Se" = "#82a1c1")

sCST_colors2 <- c("High aureus" = "#e68502", "Mid aureus" = "#f2ad00", "Low aureus" = "#fad77b", "High capitis" = "#644b63", "Lugdunensis" = "#a7bf90", "Mixed" = "#385338", "Low capitis" = "#ac8cae", "Epidermidis" = "#82a1c1")

sCST_colors3 <- c("sCST1Sa_H (n=175)" = "#e68502", "sCST1Sa_M (n=238)" = "#f2ad00", "sCST2Sa (n=180)" = "#fad77b", "sCST2Sc_H (n=61)" = "#644b63", "sCST2Sl (n=143)" = "#a7bf90" , "sCST2Sh (n=339)" = "#385338", "sCST2Sc (n=136)" = "#ac8cae", "sCST2Se (n=591)" = "#82a1c1")

CST_colors <- c("CST1a" = "#e68502", "CST1b" = "#f2ad00", "CST2" = "#713030", "CST3" = "#82a1c1", "CST4" = "#d4a1b2", "CST5a" = "#aa4365", "CST5b" = "#cd6677", "CST6a"= "#636363", "CST6b" = "#3b3b3b","CST7" = "#66283C", "CST8" = "#494978", "CST9a"= "#018855", "CST9b" = "#7ab898")

Staph_spec_colors <- c("S. aureus" = "#f2ad00", "S. epidermidis" = "#82a1c1", "S. lugdunensis" = "#a7bf90", "S. capitis" = "#644b63", "S. hominis" = "#385338", "S. warneri" = "#965a3c", "S. haemolyticus" = "#713030"  , "S. pasteuri" = "#cd6677", "S. saccharolyticus" = "#F5A078", "S. equorum" = "#dfc38d",  "Others" = "grey")

Cohort_colors <- c("Germany" = "#7896b4", "Denmark" = "#c76767")
```



## Plot frequency data (cross-sectional data only)  
```{r fig.height=8, fig.width=7}
frequency_table_top20 = frequency_table[match(row.names(species_summary[order(species_summary$sample_presence,decreasing = TRUE),])[1:20], row.names(frequency_table)),]

frequency_melt = melt(as.matrix(frequency_table_top20))
names(frequency_melt)[1] = "Species"

frequency_melt$Species = factor(frequency_melt$Species, levels = row.names(species_summary[order(species_summary$sample_presence,decreasing = TRUE),])[20:1])

levels(frequency_melt$Species) = gsub("Staphylococcus_", "S. ", levels(frequency_melt$Species))

frequency_plot <- ggplot(frequency_melt) +
  aes(x = value,
      y = Species,
      color = value) +
  geom_point(position = position_jitter(w = .15, h=0.2), alpha=0.9) +
  geom_violin(scale = "width", fill=NA,draw_quantiles = c(0.5)) +
  scale_color_gradient(low = "#fff1f1", high = "#980505",name = "Relative abundance, %")+
   scale_x_continuous(expand = c(0,1)) +
  theme_cowplot() +
  ylab("") + xlab("")+
  theme(axis.text.y = element_text(face = "italic"))+
  theme(legend.position = "bottom") 

frequency_plot
```



## Jensen-Shannon distance matrix and NMDS analysis by population on cross-sectional data only:  

Calculate distance matrix using Jensen-Shannon divergence  
```{r}
#Extract sample names of cross-sectional samples
CS <- count_metadata %>% rownames_to_column("Sample") %>% filter(data_type == "cross sectional") %>% pull(Sample)

#Subset count_table to cross-sectional samples
count_table_CS <- t(count_table) %>% as.data.frame() %>% rownames_to_column("Sample") %>% filter(Sample %in% CS) %>% column_to_rownames("Sample")

#Calculate distance matrix of cross-sectional samples
count_dis_CS = philentropy::distance(count_table_CS,method = "jensen-shannon", use.row.names = TRUE)
```

Run NMDS using the calculated distance matrix (Jensen-Shannon divergence).  
This includes only cross-sectional data.  
```{r}
set.seed(1)
count_MDS_CS = metaMDS(comm = count_dis_CS, trymax = 40)
```


### Stress plot 
We can visualise the stress of the NMDS to see if it is transformed well. 
```{r}
count_MDS_CS
stressplot(count_MDS_CS)
```


## Plot NMDS according to population (Germany vs Denmark)  
```{r}
NMDS_plot_CS = data.frame(NMDS1 = count_MDS_CS$points[,1], NMDS2 = count_MDS_CS$points[,2])

NMDS_plot_CS <- left_join(NMDS_plot_CS %>% rownames_to_column("Sample"), count_metadata %>% filter(data_type == "cross sectional") %>% rownames_to_column("Sample") %>% select(Sample, population), by = "Sample") %>% column_to_rownames("Sample")

population_nmds_CS = ggplot(data = NMDS_plot_CS, aes(x = NMDS1, y = NMDS2, color = population)) +
  geom_point(size = 1, alpha = 0.5) +
  stat_ellipse(type = "t", level = 0.95) +
  labs(color = "Population") +
  scale_color_manual(values = Cohort_colors)+
  theme_cowplot() +
  annotate("text", x = -30, y = -20, label = "Stress = 0.1") +
  theme(legend.position=c(0.02, 0.87))
population_nmds_CS
```

### Test with ANOSIM whether there is a significant difference in composition between the two study populations.  
This only includes cross-sectional data.  
```{r}
count_metadata_CS <- count_metadata %>% filter(data_type == "cross sectional")

set.seed(1)
ano_jsd_CS = anosim(count_dis_CS, count_metadata_CS$population, permutations = 1000, parallel = 7)
ano_jsd_CS
```
There is no significant difference in composition between the two study populations.  
  

## Calculate alpha diversity based on several indices and compare between populations (DK vs GER, cross-sectional data only)    

```{r}
count_metadata_CS <- count_metadata %>% filter(data_type == "cross sectional")

count_table_CS <- count_table[,colnames(count_table) %in% rownames(count_metadata_CS)]

count_diversity_CS = data.frame(row.names = row.names(count_metadata_CS),
                             population = count_metadata_CS$population)

count_diversity_CS$Richness = vegan::specnumber(t(count_table_CS))

count_diversity_CS$Shannon = vegan::diversity(t(count_table_CS), index = "shannon")

count_diversity_CS$`Inverse Simpson` = vegan::diversity(t(count_table_CS), index = "invsimpson")

diversity_melt_CS = melt(count_diversity_CS)

population_diversity_CS = ggplot(diversity_melt_CS, aes(x = population, y = value)) +
  geom_boxplot(aes(fill = population), alpha = 0.8, outlier.alpha = 1, position = position_dodge2(width = 0.5)) +
  scale_fill_manual(values=Cohort_colors)+
  ylab("Alpha diversity measurement") +
  facet_wrap(~variable,scales = "free_y") +
  theme_cowplot() +
  theme(legend.position = "", axis.title.x = element_blank(), axis.text.x = element_text(angle = 40, vjust = 0.6))
population_diversity_CS
```



### Test alpha diversity measures between DK and GER (Kruskal Wallis tests), cross-sectional data only  
```{r}
count_diversity_CS %>% dplyr::summarise(p_Richness = kruskal.test(Richness~population)$p.value, 
                    p_Shannon = kruskal.test(Shannon~population)$p.value,
                    p_InvSimpson = kruskal.test(`Inverse Simpson`~ population)$p.value)
```


# Combine plot 
```{r fig.height=4, fig.width=10}
population_grid2_CS20 <-   
  plot_grid(population_nmds_CS, population_diversity_CS,labels = c("A", "B"),label_size = 18,  nrow = 1,rel_widths = c(1,1.35))

population_grid2_CS20
```



## S. aureus prevalence and abundance in GER vs DK
```{r}
df_aureus <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(data_type == "cross sectional") 
```

```{r}
prev_table <- df_aureus %>% group_by(population) %>% 
  dplyr::summarise(S.aureus_absent_n = matrixStats::count(Staphylococcus_aureus == 0) , S.aureus_present_n= matrixStats::count(Staphylococcus_aureus>0)) %>% 
 group_by(population)  %>% 
  dplyr::summarise("S.aureus_present >0%" = paste0(S.aureus_present_n, " (", round(S.aureus_present_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)"),
            "S.aureus_absent (0%)" = paste0(S.aureus_absent_n, " (", round(S.aureus_absent_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)")) 

prev_table %>%
  kable() %>% kable_classic()
```

```{r}
prev_table10 <- df_aureus %>% group_by(population) %>% 
  dplyr::summarise(S.aureus_absent_n = matrixStats::count(Staphylococcus_aureus <10) , S.aureus_present_n= matrixStats::count(Staphylococcus_aureus>=10)) %>% 
 group_by(population)  %>% 
  dplyr::summarise("S.aureus_present >=10%" = paste0(S.aureus_present_n, " (", round(S.aureus_present_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)"),
            "S.aureus_absent (<10%)" = paste0(S.aureus_absent_n, " (", round(S.aureus_absent_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)")) 

prev_table10 %>%
  kable() %>% kable_classic()
```

Which abundance would be required to detect 43% of samples as positive? (corresponding to culture-positive prevalence) 
```{r}
prev_table04 <- df_aureus %>% group_by(population) %>% 
    dplyr::summarise(S.aureus_absent_n = matrixStats::count(Staphylococcus_aureus <0.4) , S.aureus_present_n= matrixStats::count(Staphylococcus_aureus>=0.4)) %>% 
    group_by(population)  %>% 
   dplyr::summarise("S.aureus_present >=10%" = paste0(S.aureus_present_n, " (", round(S.aureus_present_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)"),
            "S.aureus_absent (<10%)" = paste0(S.aureus_absent_n, " (", round(S.aureus_absent_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)")) 

prev_table04 %>%
  kable() %>% kable_classic()
```

```{r}
df_aureus2 <- as.data.frame(t(count_table[,colnames(count_table) %in% df_aureus$Sample])) %>% select(-Staphylococcus_aureus) %>%
  rownames_to_column("Sample") %>% 
  rowwise(Sample) %>%  dplyr::summarise(Other_Staph=sum(c_across(where(is.numeric))))  %>% 
  left_join(df_aureus) %>% 
  mutate(Staph_aureus = Staphylococcus_aureus) %>% 
  pivot_longer(cols= c(Other_Staph, Staphylococcus_aureus))
```

```{r fig.height=6, fig.width=8}
df_aureus2_GER <- df_aureus2 %>% filter(population == "Germany") %>% arrange(Staph_aureus)
df_aureus2_DK <- df_aureus2 %>% filter(population == "Denmark") %>% arrange(Staph_aureus)


GER <- ggplot(df_aureus2_GER, aes(x=Sample, y=value, fill=name)) +
geom_bar(position = "stack", stat = "identity", width = 1) +
scale_fill_manual(values=c("grey", "#1f78b4"), labels=c("Other staphylococci", "S. aureus")) +
scale_x_discrete(limits=df_aureus2_GER$Sample) +
theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = c(0.1, 0.75), legend.title = element_blank()) + 
labs(y = "Relative abundance [%]", x = "Study participants") + ggtitle("Germany") + 
geom_hline(yintercept = 10, color="red", size=0.5) + 
geom_vline(xintercept = df_aureus2_GER$Sample[df_aureus2_GER$Staph_aureus >0][1], size=0.5) + 
geom_vline(xintercept = df_aureus2_GER$Sample[df_aureus2_GER$Staph_aureus >=10][1], size=0.5) + 
annotate(x= df_aureus2_GER$Sample[df_aureus2_GER$Staph_aureus >0][120], y= 90, label=paste0("S. aureus present >0%: \n", "n=", prev_table$`S.aureus_present >0%`[prev_table$population == "Germany"]), "text") +
annotate(x= df_aureus2_GER$Sample[df_aureus2_GER$Staph_aureus >=10][125], y= 90, label=paste0("S. aureus present >=10%: \n", "n=", prev_table10$`S.aureus_present >=10%`[prev_table10$population == "Germany"]), "text") +  
scale_y_continuous(expand = c(0,0)) 


DK <- ggplot(df_aureus2_DK, aes(x=Sample, y=value, fill=name)) +
geom_bar(position = "stack", stat = "identity", width = 1) +
scale_fill_manual(values=c("grey", "#1f78b4")) +
scale_x_discrete(limits=df_aureus2_DK$Sample) +  
theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + 
labs(y = "Relative abundance [%]", x = "Study participants") + ggtitle("Denmark")+ 
geom_hline(yintercept = 10, color="red", size=0.5)+ 
geom_vline(xintercept = df_aureus2_DK$Sample[df_aureus2_DK$Staph_aureus >0][1], size=0.5) + 
geom_vline(xintercept = df_aureus2_DK$Sample[df_aureus2_DK$Staph_aureus >=10][1], size=0.5) + 
annotate(x= df_aureus2_DK$Sample[df_aureus2_DK$Staph_aureus >0][200], y= 90, label=paste0("S. aureus present >0%: \n", "n=", prev_table$`S.aureus_present >0%`[prev_table$population == "Denmark"]), "text") +
annotate(x= df_aureus2_DK$Sample[df_aureus2_DK$Staph_aureus >=10][220], y= 90, label=paste0("S. aureus present >=10%: \n", "n=", prev_table10$`S.aureus_present >=10%`[prev_table10$population == "Denmark"]), "text") +  
scale_y_continuous(expand = c(0,0)) 

SA_pres <- cowplot::plot_grid(DK, GER, nrow = 2)
SA_pres
```


How many of the Danish samples with S. aureus rel abund >=0 at t1, had >=0 at t2 or t2 and t3 (recurrence percentage)
```{r}
DK_1_pos <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(population == "Denmark", timepoint == "timepoint1", Staphylococcus_aureus>0) %>% pull(Participant_ID) 
```

How many of the positive ones have a 2nd time point? And how many of these are still positive?

```{r}
DK_2_pos <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(timepoint == "timepoint2", Participant_ID %in% DK_1_pos)

table(DK_2_pos$Staphylococcus_aureus>0)
```


```{r}
112/159 #recurrence percentage (carrier at two time points) 70%
#That corresponds to
0.7*75 #53% of general carriers being persistent carriers
#52.8% If 159 is 75% (carriers at tp1) then 212 is 100%, i.e. 112 (persistent carriers) is 53% of the entire cohort
112/212
```


How many have both tp 2 and tp 3 and stay positive at all three, i.e. actually persistent


```{r}
DK_23_pos <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(population == "Denmark", Participant_ID %in% DK_1_pos) %>% pivot_wider(id_cols= Participant_ID, names_from = timepoint, values_from = Staphylococcus_aureus) %>% filter(!is.na(timepoint2)) %>% filter(!is.na(timepoint3))

dim(DK_23_pos)
table(DK_23_pos$timepoint2>0, DK_23_pos$timepoint3>0)
```
```{r}
28/59 #recurrence percentage (carrier at all three time points) 47% of tp1 carriers are also carriers at both tp 2 and tp 3 
#That correponds to 
0.47*75 #35% of general carriers being persistent carriers or 28/((59/75)*100)
```


How many of the negative ones have a 2nd time point? And how many of these are still negative?

```{r}
DK_1_neg <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(population == "Denmark", timepoint == "timepoint1", Staphylococcus_aureus==0) %>% pull(Participant_ID) 

#length(DK_1_neg)

DK_2_neg <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(timepoint == "timepoint2", Participant_ID %in% DK_1_neg)

table(DK_2_neg$Staphylococcus_aureus==0)
```

58 out of the 232 have a a sample at tp2, 19 of these are still negative, i.e. 19/58 = 33%. 


How many have both tp 2 and tp 3 and stay negative at all three, i.e. actually non-carriers
```{r}
DK_23_neg <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(population == "Denmark", Participant_ID %in% DK_1_neg) %>% pivot_wider(id_cols= Participant_ID, names_from = timepoint, values_from = Staphylococcus_aureus) %>% filter(!is.na(timepoint2)) %>% filter(!is.na(timepoint3))

#dim(DK_23_neg)
table(DK_23_neg$timepoint2==0, DK_23_neg$timepoint3==0)
```
20 participants who are neg at tp1 have 2 follow-up samples. 4 out of 20 are negative at all three time points (20%). 

Visualize:

```{r fig.height=10, fig.width=5}
Dk_neg_df <- as.data.frame(t(count_table)) %>% select(Staphylococcus_aureus) %>% 
  rownames_to_column("Sample") %>% left_join(count_metadata %>% rownames_to_column("Sample")) %>% 
  filter(population == "Denmark") %>% 
  pivot_wider(id_cols= Participant_ID, names_from = timepoint, values_from = Staphylococcus_aureus) %>% 
  filter(!is.na(timepoint1), !is.na(timepoint2), !is.na(timepoint3)) %>% 
  mutate(tuf_carrier = ifelse(timepoint1 == 0 & timepoint2 == 0 & timepoint3 == 0, "1. non-carrier", ifelse(timepoint1 > 0 & timepoint2 > 0 & timepoint3 > 0, "3. persistent carrier", 
        ifelse(timepoint1 > 0 & timepoint2 > 0 & timepoint3 ==0,"2.4 intermittent carrier (lost)",
         ifelse(timepoint1 > 0 & timepoint2 ==0 & timepoint3 ==0,"2.3 intermittent carrier (lost)",                 ifelse(timepoint1 ==0 & timepoint2 ==0 & timepoint3 >0,"2.1 intermittent carrier (gained)",
         ifelse(timepoint1 ==0 & timepoint2 >0 & timepoint3 >0,"2 intermittent carrier (gained)",
        ifelse(timepoint1 ==0 & timepoint2 >0 & timepoint3 ==0,"2.2 intermittent carrier (gained and lost)", "2.5 intermittent carrier (lost and gained)")))))))) %>%
pivot_longer(names_to = "timepoint", values_to = "Staphylococcus_aureus", cols = c(timepoint1, timepoint2, timepoint3))

ggplot(Dk_neg_df, aes(y=Participant_ID,x=timepoint, fill=Staphylococcus_aureus>0)) +
  geom_tile(width=0.8, height=0.7)+
  facet_grid(tuf_carrier~., scales = "free", space= "free") +
  theme(axis.text.y=element_text(size=8), axis.title.x = element_blank(), axis.title.y = element_blank(), legend.position = "top", strip.text.y = element_text(angle = 0)) +
  scale_fill_manual(values = c("grey", "#e8702a"))+
  labs(fill="S. aureus carriage (tuf)")
```




# Pairwise correlations of staphylococcal species (relative abundance) in cross-sectional samples    
Here we analyse which staphylococcal species correlate positively or negatively with each other according to their relative abundances.  
Only correlations >0.1 or <-0.1 are displayed.  
```{r fig.height=6, fig.width=6}
count_table_t <- as.data.frame(t(count_table))

#Subset cross-sectional samples only: 
CS <- count_metadata %>% rownames_to_column("Sample") %>% filter(data_type == "cross sectional") %>% pull(Sample)

count_table_t_CS <- count_table_t %>% rownames_to_column("Sample") %>% filter(Sample %in% CS)
dim(count_table_t_CS)

count_table_t_CS_sub <- count_table_t_CS %>% select(Staphylococcus_aureus, Staphylococcus_epidermidis, Staphylococcus_capitis, Staphylococcus_lugdunensis, Staphylococcus_hominis, Staphylococcus_warneri, Staphylococcus_haemolyticus, Staphylococcus_pasteuri, Staphylococcus_saccharolyticus, Staphylococcus_equorum)

M = cor(count_table_t_CS_sub, method= "spearman") 

#Subset correlations >0.1/<-0.1:

#matrix
M2 <- M %>% as.data.frame() %>% mutate(across(everything(), ~ifelse(. >=0.1, ., ifelse(. <=-0.1, ., 0)))) %>% as.matrix()

#dataframe
M_df <- M %>% as.data.frame() %>% mutate(across(everything(), ~ifelse(. >=0.1, ., ifelse(. <=-0.1, ., 0)))) %>%
    rownames_to_column("Staph1") %>% pivot_longer(cols= c(2:ncol(.)), names_to = "Staph2", values_to = "rho")
```


Remove self-correlations (set to 0)
```{r}
#matrix
M3 <- M2 %>% as.data.frame() %>% mutate(across(everything(), ~ifelse(. ==1, 0, .))) %>% 
  select( -Staphylococcus_saccharolyticus, -Staphylococcus_equorum) %>% 
  filter(rownames(.) != "Staphylococcus_aureus") %>% 
  as.matrix()

#dataframe
M_df2 <- M2 %>% as.data.frame() %>% mutate(across(everything(), ~ifelse(. ==1, 0, .))) %>%
    select( -Staphylococcus_saccharolyticus, -Staphylococcus_equorum)   %>% 
    filter(rownames(.) != "Staphylococcus_aureus") %>% 
    rownames_to_column("Staph1") %>% pivot_longer(cols= c(2:ncol(.)), names_to = "Staph2", values_to = "rho")
```


Calculate significance of the correlations  
```{r warning=FALSE}
# make significance matrix
testRes = cor.mtest(count_table_t_CS_sub, conf.level = 0.95, method = "spearman", exact=FALSE)

testRes_sign <- data.frame(testRes$p) %>%
  mutate(across(.cols=everything(),p.adjust,method="BH"))

#Subset correlations >0.1/<-0.1
testRes_sign2 <- testRes_sign %>% rownames_to_column("Staph1") %>% pivot_longer(cols= c(2:ncol(.)), names_to = "Staph2", values_to = "padj") %>% 
  left_join(M_df2) %>% mutate(padj = ifelse(rho == 0, 1, padj)) %>% select(-rho) %>% pivot_wider(names_from = "Staph2", values_from = "padj") %>%   
  column_to_rownames("Staph1") %>% 
   select( -Staphylococcus_saccharolyticus, -Staphylococcus_equorum)  %>% filter(rownames(.) != "Staphylococcus_aureus") 
```


### Correlation plot
```{r fig.height=6, fig.width=9}
corrplot(M3, method ="color", type="lower",  is.corr = FALSE,
         p.mat=as.matrix(testRes_sign2),sig.level = c(0.001), pch.cex = 0.65,insig = 'label_sig', pch = c("p<0.001"), pch.col = 'black', col.lim= c(-1,1),col= colorRampPalette(c("#003C30", "white", "#8E0152"))(11), addgrid.col = 'grey')

```




# Hierarchical clustering of samples into staphylococcal community types (sCSTs)  

### Calculate distance matrix using Jensen-Shannon divergence (JSD)    
```{r}
count_dis = philentropy::distance(t(count_table),method = "jensen-shannon",use.row.names = TRUE)
```

### Determine the appropriate amount of clusters  
Three methods are used: Within-Cluster Sum of Squares (wss), silhouette method, and gap statistics.  
```{r}
pcoa_count = ape::pcoa(D = count_dis)
clust_wss = fviz_nbclust(pcoa_count$vectors, hcut, method = "wss")
clust_sil = fviz_nbclust(pcoa_count$vectors, hcut, method = "silhouette")
clust_gap = fviz_nbclust(pcoa_count$vectors, hcut, method = "gap_stat",k.max = 10,nboot = 50)
```

### Plot evaluation of optimal number of clusters  
```{r fig.height=8, fig.width=5}
cluster_grid = plot_grid(clust_wss, clust_sil, clust_gap, align = "hv", labels = "AUTO", ncol = 1)
cluster_grid
```


### Cluster the samples using hierachical clustering
The appropriate amount of clusters is 2 to 3. We determined that there is 2 main groups which are further subdivided into 8 distinct subgroups.  
```{r}
count_hc = hclust(as.dist(count_dis), method = "ward.D")
plot(count_hc, cex = 0.5, labels = F, hang = 0)
rect.hclust(count_hc, k = 8)
count_metadata$cutGroup = cutree(count_hc, k = 8)
count_metadata$cutGroup = factor(count_metadata$cutGroup, levels = c(1:8))
```

Create a version of the count table where the top 10 most abundant species are selected, and the remaining species are summarized into "Others".  
```{r}
count_table2 = count_table
g_list = names(sort(rowSums(count_table), decreasing = TRUE)[1:10])
g_list_other = names(sort(rowSums(count_table), decreasing = TRUE)[11:40])
g_sum = colSums(count_table2[(match(g_list_other, row.names(count_table2))),])
count_table2 = count_table2[match(g_list,row.names(count_table2)),]
count_table2 = rbind(count_table2,g_sum)
row.names(count_table2)[11] = "Others"
```

### Initial barplot of the species
This initial figure plots directly the group allocated by the hierarchical clustering.  
```{r fig.width=12}
count_melt = melt(t(count_table2))
colnames(count_melt)[1] = "Sample"
colnames(count_melt)[2] = "Species"
colnames(count_melt)[3] = "Count"
count_melt$cutGroup = count_metadata$cutGroup[match(count_melt$Sample, row.names(count_metadata))]

count_melt = within(count_melt, Sample <- factor(Sample, levels = as.character(unique(count_melt$Sample[order(count_melt$cutGroup)]))))

levels(count_melt$Species) = gsub("Staphylococcus_", "S. ", levels(count_melt$Species))

ggplot(count_melt, aes(x = Sample, y = Count, fill = Species)) +
  geom_bar(position = "fill", stat = "identity", width = 1) +
  ylab("Relative abundance") +
  scale_fill_manual(values=Staph_spec_colors) + 
  facet_wrap(~ cutGroup, ncol = 4,scales = "free_x") +
  theme_cowplot() +
  scale_y_continuous(expand = c(0,0), labels = scales::label_percent()) +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x=element_blank()) +
  theme(legend.position = "bottom")
```

#### Rename each group into the appropriate sCST based on the observed distribution of staphylococcal species.  
The groups are given names. They are characterized as follows:   
  
* sCST1 - *S. aureus* dominant    
  + High aureus - sCST1~*Sa*_H~   
  + Mid aureus - sCST1~*Sa*_M~    
* sCST2 - *S. epidermidis* dominant   
  + Low aureus - sCST2~*Sa*~    
  + High capitis - sCST2~*Sc*_H~    
  + Lugdunensis - sCST2~*Sl*~   
  + Mixed - sCST2~*Sh*~    
  + Low capitis - sCST2~*Sc*~   
  + Epidermidis - sCST2~*Se*~   
  
The sCST assignments are written into the metadata table.   
The table below shows the number (n) and percentage of samples groupd into each sCST.  
```{r}
count_metadata$sCST = ""
count_metadata$sCST[which(count_metadata$cutGroup==1)] = "High aureus"
count_metadata$sCST[which(count_metadata$cutGroup==2)] = "Epidermidis"
count_metadata$sCST[which(count_metadata$cutGroup==3)] = "Mid aureus"
count_metadata$sCST[which(count_metadata$cutGroup==4)] = "Low capitis"
count_metadata$sCST[which(count_metadata$cutGroup==5)] = "Lugdunensis"
count_metadata$sCST[which(count_metadata$cutGroup==6)] = "Low aureus"
count_metadata$sCST[which(count_metadata$cutGroup==7)] = "High capitis"
count_metadata$sCST[which(count_metadata$cutGroup==8)] = "Mixed"
count_metadata$sCST = factor(count_metadata$sCST, levels = c("High aureus", "Mid aureus", "Low aureus","High capitis","Lugdunensis", "Mixed","Low capitis", "Epidermidis"))

count_metadata %>% group_by(sCST) %>% tally() %>%  mutate(percentage = round(n/sum(n)*100, digit = 1)) %>% kable() %>% kable_classic()
```

#### Transferring names into the bar plot    
```{r fig.width=12}
count_melt$sCST = ""
count_melt$sCST[which(count_melt$cutGroup==1)] = "sCST1Sa_H (n=175)"
count_melt$sCST[which(count_melt$cutGroup==2)] = "sCST2Se (n=591)"
count_melt$sCST[which(count_melt$cutGroup==3)] = "sCST1Sa_M (n=238)"
count_melt$sCST[which(count_melt$cutGroup==4)] = "sCST2Sc (n=136)"
count_melt$sCST[which(count_melt$cutGroup==5)] = "sCST2Sl (n=143)"
count_melt$sCST[which(count_melt$cutGroup==6)] = "sCST2Sa (n=180)"
count_melt$sCST[which(count_melt$cutGroup==7)] = "sCST2Sc_H (n=61)"
count_melt$sCST[which(count_melt$cutGroup==8)] = "sCST2Sh (n=339)"
count_melt$sCST = factor(count_melt$sCST, levels = unique(count_melt$sCST)[c(1,3,6,7,5,8,4,2)])

species_temp = lapply(levels(count_melt$Species), function(x) bquote(italic(.(x))))
species_temp[12] = "Others"

group_barplot = ggplot(count_melt, aes(x = Sample, y = Count, fill = Species)) +
  geom_bar(position = "fill", stat = "identity", width = 1) +
  ylab("Relative abundance") +
   scale_fill_manual(values=Staph_spec_colors) + 
  facet_wrap(~ sCST, nrow = 2, scales = "free_x") +
  theme_cowplot() +
  scale_y_continuous(expand = c(0,0), labels = scales::label_percent()) +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks = element_blank(), legend.position = "bottom", legend.text.align = 0)
group_barplot
```

#### Calculate the mean relative abundance per staphylococcal species for each sCST.  

Combine relative count data with metadata on sCST assignment.  
```{r}
count_table3 = as.data.frame(t(count_table2))
match_list = match(row.names(count_table3), row.names(count_metadata))
count_table3$sCST = as.vector(count_metadata$sCST[match_list])
count_table3$sCST = factor(count_table3$sCST, levels = c("High aureus", "Mid aureus", "Low aureus","High capitis", "Lugdunensis", "Mixed","Low capitis", "Epidermidis"))
levels(count_table3$sCST) = c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa", "sCST2Sc_H", "sCST2Sl", "sCST2Sh","sCST2Sc", "sCST2Se")
```

Apart from the 5 most dominant species, other coagulase-negative staphylococci are grouped together as "CoNS", while the remaining species are grouped as "Other_Staph".  
```{r}
temp_vec_CoNS <- count_table %>% rownames_to_column("Species") %>% filter(!Species %in% c("Staphylococcus_aureus", "Staphylococcus_epidermidis", "Staphylococcus_capitis", "Staphylococcus_lugdunensis", "Staphylococcus_delphini", "Staphylococcus_lutrae", "Staphylococcus_pseudintermedius", "Staphylococcus_schleiferi", "Staphylococcus_hominis")) %>% column_to_rownames("Species") %>% dplyr::summarise_all(.,sum) %>% pivot_longer(names_to = "Sample", values_to = "CoNS", cols = everything()) %>% column_to_rownames("Sample")

temp_vec_non_CoNS <- count_table %>% rownames_to_column("Species") %>% filter(Species %in% c( "Staphylococcus_delphini", "Staphylococcus_lutrae", "Staphylococcus_pseudintermedius", "Staphylococcus_schleiferi")) %>% column_to_rownames("Species") %>% dplyr::summarise_all(.,sum) %>% pivot_longer(names_to = "Sample", values_to = "Other_staphylococci", cols = everything()) %>% column_to_rownames("Sample")

count_table3$CoNS = temp_vec_CoNS$CoNS
count_table3$Other_Staph = temp_vec_non_CoNS$Other_staphylococci
```

### Show the mean relative abundance per staphylococcal species/group for each sCST
```{r}
count_summary = count_table3 %>% group_by(sCST) %>% dplyr::summarise(aureus = mean(Staphylococcus_aureus), epidermidis = mean(Staphylococcus_epidermidis), capitis = mean(Staphylococcus_capitis), lugdunensis = mean(Staphylococcus_lugdunensis), hominis = mean(Staphylococcus_hominis), CoNS = mean(CoNS), Other_Staph = mean(Other_Staph)) %>% mutate_at(2:8, ~(round(., digits = 1)))
count_summary %>% kable() %>% kable_classic()
```


# Ordination analysis  
  
Run NMDS using the previously calculated distance matrix (Jensen-Shannon divergence).  
This includes both cross-sectional and longitudinal data.  
```{r}
set.seed(1)
count_MDS = metaMDS(comm = count_dis, trymax = 40)
```

### Stress plot 
We can visualize the stress of the NMDS to see if it is transformed well. 
```{r}
count_MDS
stressplot(count_MDS)
```


### Plot NMDS according to the 8 sCSTs.  
Extracting the NMDS data from the object and plotting in ggplot2.
```{r}
NMDS_plot = data.frame(NMDS1 = count_MDS$points[,1], NMDS2 = count_MDS$points[,2])
NMDS_plot$sCST = count_metadata$sCST
NMDS_plot2 = ggplot(data = NMDS_plot, aes(x = NMDS1, y = NMDS2, color = sCST)) +
  geom_point(size = 1, alpha = 0.5) +
  stat_ellipse(type = "t", level = 0.95) +
  labs(color = "sCST") +
  scale_color_manual(values = sCST_colors2) +
  theme_cowplot() +
  annotate("text", x = -30, y = -20, label = "Stress = 0.1") +
  theme(legend.position = "right")
NMDS_plot2
```


# Heatmap with dendrogram of the sCSTs  
Next, we generate a heatmap based on the relative abundance of the top 10 species (and "Others"). The hierachical clustering dendrogram is used to show the sCST groupings.

#### Initial plot
```{r }
count_heatmap_column = count_metadata %>% select(population, sCST)

levels(count_heatmap_column$sCST) = c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa", "sCST2Sc_H", "sCST2Sl", "sCST2Sh","sCST2Sc", "sCST2Se")

names(count_heatmap_column)[1] = "Population"
names(count_heatmap_column)[2] = "sCST"

count_column_color = list(Population = Cohort_colors, sCST = sCST_colors)

row_temp_name = lapply(gsub("Staphylococcus_", "S. ", rownames(count_table2)), function(x) bquote(italic(.(x))))
row_temp_name[11] = "Others"

heat_count = pheatmap(mat = count_table2, color = colorRampPalette(c("#fff1f1", "#980505"))(25), cutree_cols = 8, cluster_cols = count_hc, cluster_rows = FALSE, show_colnames = FALSE, annotation_col = count_heatmap_column, annotation_colors = count_column_color, treeheight_col = 100, labels_row = as.expression(row_temp_name), fontsize=7)
```

## Reorder the dendrogram
Here we rotate the dendrogram so that sCST1 (S. aureus) is on the left and sCST2 (S. epidermidis) on the right.  
```{r }
count_hc2 = rotate(count_hc, order = c(1451:1863,1271:1450,1270:1))
heat_count = pheatmap(mat = count_table2, color = colorRampPalette(c("#fff1f1", "#980505"))(25), cutree_cols = 8, cluster_cols = count_hc2, cluster_rows = FALSE, show_colnames = FALSE, annotation_col = count_heatmap_column, annotation_colors = count_column_color, treeheight_col = 100, labels_row = as.expression(row_temp_name), fontsize=7)
heat_count
```






Combine dendrogram and stacked barplot (horizontally)
```{r fig.height=20, fig.width=10}
tree <- dendro_data(count_hc2)

data <- cbind(count_melt, x = match(count_melt$Sample, tree$labels$label))
data <- data %>% left_join(count_metadata %>% select(population) %>% rownames_to_column("Sample"), by="Sample")

colZ <- c(Staph_spec_colors,sCST_colors3, Cohort_colors)
```


```{r fig.height=8, fig.width=18}
scale <- 0.005
ph <- ggplot() +
  
 geom_segment(
    data = tree$segments,
    aes(x = x, y = (-y *scale) -10, xend = xend, yend = (-yend*scale) -10)) + 
  
  geom_col(
    data = data,
    aes(x = x,
        y = Count, fill = Species))  +
  
  scale_y_reverse()+
  
  geom_tile(data=data, aes(x = x, y = -5, fill = sCST, height = 8))+
 
 scale_fill_manual(values=colZ)+
  
 geom_text(aes(x=80,y=-5,label=c("sCST1Sa_H (n=175)")), size = 4)+
 geom_text(aes(x=300,y=-5,label=c("sCST1Sa_M (n=238)")), size = 4)+
 geom_text(aes(x=500,y=-5,label=c("sCST2Sa (n=180)")), size = 4)+
 geom_text(aes(x=650,y=-5,label=c("sCST2Sc_H \n(n=61)")), size = 4, color="white")+
 geom_text(aes(x=750,y=-5,label=c("sCST2Sl (n=143)")), size = 4)+
 geom_text(aes(x=1000,y=-5,label=c("sCST2Sh (n=339)")), size = 4, color="white")+
 geom_text(aes(x=1200,y=-5,label=c("sCST2Sc \n(n=136)")), size = 4)+
 geom_text(aes(x=1550,y=-5,label=c("sCST2Se (n=591)")), size = 4)+
  

  theme_cowplot()+
  theme(axis.line.x=element_blank(),axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "top") +
  labs(x="", y= "Relative abundance [%]")

ph
```



# Alpha diversity  

### Calculate Shannon diversity across different sCSTs
```{r}
count_diversity = data.frame(row.names = row.names(count_metadata),
                             sCST = count_metadata$sCST)

count_diversity$shannon = vegan::diversity(t(count_table), index = "shannon")

levels(count_diversity$sCST) = c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa", "sCST2Sc_H","sCST2Sl", "sCST2Sh", "sCST2Sc", "sCST2Se")

shannon_diversity = ggplot(data = count_diversity, aes(x = sCST, y = shannon, fill = sCST)) +
  geom_violin(scale = "width") +
  geom_boxplot(width = 0.2,outlier.alpha = 0) +
  scale_fill_manual(values=sCST_colors) + 
  scale_y_continuous(expand = c(0,0)) +
  theme_cowplot() +
  ylab("Shannon diversity") +
  xlab("") +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 9)) +
  theme(legend.position = "", axis.text.x = element_text(angle=30, hjust=1))
shannon_diversity
```

### Test alpha diversity measures between sCSTs.  
This is not trivial, because some data is paired since the longitudinal samples are included. However, let's do a pairwise wilcoxon test as an approximation even though it's not 100% appropriate (therefore result not used in manuscript).    
```{r}
pairwise.wilcox.test(count_diversity$shannon, count_diversity$sCST, p.adjust.method = "BH")
```


Combine plot  
```{r fig.height=10, fig.width=10}
F4 = plot_grid(NMDS_plot2 + theme(legend.position = "none"), shannon_diversity, labels = c("B","C"),align = "hv",rel_widths = c(3:5),label_size = 19)
#F4.1 = plot_grid (as.ggplot(heat_count), NULL, F4, group_barplot, ncol = 1, rel_heights = c(9,0.15,8,10), labels = c("A","","","D"),label_size = 19)
F4.1 = plot_grid (ph, NULL, F4, ncol = 1, rel_heights = c(12,0.15,8,10), labels = c("A","",""),label_size = 19)
F4.1
```



# Longitudinal data analysis  

Format data (DK only)
```{r}
count_metadata_longi <- count_metadata %>% filter(population == "Denmark") %>% pivot_wider(id_cols= Participant_ID, names_from = timepoint, values_from = sCST)

# remove participants that don't have the 2nd time point
count_metadata_longi <- count_metadata_longi %>% filter(!is.na(timepoint2))

dim(count_metadata_longi)
```

### Create an alluvial plot showing sCST changes over time for the 79 participants that have data at 3 timepoints.
```{r}
timepoint3 = count_metadata_longi
timepoint3 = na.exclude(timepoint3)
timepoint3_melt = melt(timepoint3, id.vars = "Participant_ID")
timepoint3_melt$value = as.vector(timepoint3_melt$value)
timepoint3_melt$value = factor(timepoint3_melt$value, levels = c("High aureus", "Mid aureus","Low aureus", "High capitis","Lugdunensis", "Mixed", "Low capitis" ,"Epidermidis"))
levels(timepoint3_melt$value) = c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa", "sCST2Sc_H", "sCST2Sl", "sCST2Sh","sCST2Sc", "sCST2Se")
timepoint3_melt$freq = 1

time3 = ggplot(timepoint3_melt, aes(x = variable, stratum = value, alluvium = Participant_ID, label = freq, fill = value)) +
  geom_flow(stat = "alluvium") +
  scale_y_continuous(name = "Number of individuals", expand = c(0,0)) +
  scale_x_discrete(name = "Time point",expand = c(0,0.25), labels = c("sCST1" = "1", "sCST2"= "2", "sCST3" = "3")) +
  geom_stratum() +
  scale_fill_manual(values = sCST_colors) + 
  labs(fill = "sCST") +
  theme_cowplot() +
  theme(legend.position = "bottom")
time3
```



## Markov chains  
Markov chain models are used to calculate transition probabilities over time between sCSTs.  


Prepare data   
```{r}
temp_timepoint3 = data.frame(timepoint1 = as.vector(count_metadata_longi$timepoint1),
                             timepoint2 = as.vector(count_metadata_longi$timepoint2),
                             timepoint3 = as.vector(count_metadata_longi$timepoint3))

# adjust naming of sCSTs:
temp_timepoint4 = apply(temp_timepoint3,2 ,function(x) sub("High aureus", "sCST1Sa_H", x))
temp_timepoint4 = apply(temp_timepoint4,2 ,function(x) sub("Mid aureus", "sCST1Sa_M", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("Low aureus", "sCST2Sa", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("Epidermidis", "sCST2Se", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("Mixed", "sCST2Sh", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("Lugdunensis", "sCST2Sl", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("High capitis", "sCST2Sc_H", x))
temp_timepoint4 = apply(temp_timepoint4,2, function(x) sub("Low capitis", "sCST2Sc", x))

#Remove NAs so we only have those 79 persons with all three time points available:
temp_timepoint4 <- as.data.frame(temp_timepoint4)
temp_timepoint5 <- temp_timepoint4 %>% filter(!is.na(timepoint1), !is.na(timepoint2), !is.na(timepoint3))
```

### Visualize transitions between the two major sCSTs (sCST1 and sCST2)  
```{r fig.height=4.5, fig.width=5}
#Make a specific dataframe that combines the sCSTs subgroups into the two major sCSTs (sCST1 and sCST2) 
temp_timepoint5_B <- temp_timepoint5 %>% 
   mutate(timepoint1 = ifelse(timepoint1 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              "sCST2")) %>% 
   mutate(timepoint2 = ifelse(timepoint2 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              "sCST2")) %>% 
   mutate(timepoint3 = ifelse(timepoint3 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              "sCST2")) 
  
temp_timepoint5_B <- as.matrix(temp_timepoint5_B)

chain_B = markovchainFit(temp_timepoint5_B) # This calculates the average of tp1 to tp2 and tp2 to tp3 transitions, which is what we want

markov_12 = data.frame(chain_B$estimate@transitionMatrix)
markov_12_round = round(markov_12, digits = 2)

#Table of transition probabilities
markov_12_round %>% kable() %>% kable_classic()

#Plot
plotmat(as.matrix(t(markov_12_round)), pos = c(2), lwd = 1.5, self.lwd = 1.5,box.lwd = 1, cex.txt = 0.7, box.size = 0.1, box.type = "circle", box.prop = 1, box.col = c("#e68502", "#82a1c1"), arr.length = 0.15, arr.width = 0.15, self.cex = 0.54, self.shifty = c(0,0), self.shiftx = c(-0.14,0.14), shadow.size=0, arr.type = "triangle")
```


### Visualize transitions between sCST1, sCST2_Sa and other sCST2
```{r fig.height=5, fig.width=5}
#Make a specific df that combines the sCSTs into those larger categories
temp_timepoint5_C <- temp_timepoint5 %>% 
   mutate(timepoint1 = ifelse(timepoint1 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              ifelse(timepoint1 == "sCST2Sa", timepoint1,
                                     "xOther_sCST2"))) %>% 
   mutate(timepoint2 = ifelse(timepoint2 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              ifelse(timepoint2 == "sCST2Sa", timepoint2,
                                     "xOther_sCST2"))) %>% 
   mutate(timepoint3 = ifelse(timepoint3 %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", 
                              ifelse(timepoint3 == "sCST2Sa", timepoint3,
                                     "xOther_sCST2"))) 
  
temp_timepoint5_C <- as.matrix(temp_timepoint5_C)

chain_C = markovchainFit(temp_timepoint5_C) # This calculates the average of tp1 to tp2 and tp2 to tp3 transitions, which is what we want

markov_aureus = data.frame(chain_C$estimate@transitionMatrix)
markov_aureus_round = round(markov_aureus, digits = 2)
markov_aureus_round <- markov_aureus_round  %>% 
  dplyr::rename("Other \nsCST2" = "xOther_sCST2")

#Table of transition probabilities
markov_aureus_round %>% kable() %>% kable_classic()

#Plot
plotmat(as.matrix(t(markov_aureus_round)), pos = c(1,2), lwd = 1.5, self.lwd= 1.5, box.lwd = 1, cex.txt = 0.7, box.size = 0.1, box.type = "circle", box.prop = 1, box.col = c("#e68502","#fad77b", "#82a1c1"),  arr.length = 0.15, arr.width = 0.15, self.cex = 0.54, self.shifty = c(0.14,0,0), self.shiftx = c(0,-0.14,0.14), shadow.size=0, arr.type = "triangle")
```



### Visualize transitions for S.aureus (sCST1Sa_H, sCST1Sa_M, sCST2Sa) vs sCST2Sl vs sCST2Se vs capitis/mixed (sCST2Sc_H, sCST2Sc, sCST2Sh)
```{r fig.height=6, fig.width=6}
#Make a specific df that combines the sCSTs into those larger categories
temp_timepoint5_D <- temp_timepoint5 %>% 
   mutate(timepoint1 = ifelse(timepoint1 %in% c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa"), "sCST1Sa_H_sCST1Sa_M_sCST2Sa", 
                              ifelse(timepoint1 %in% c("sCST2Sc_H", "sCST2Sc", "sCST2Sh"), "sCST2Sc_H_sCST2Sc_sCST2Sh",
                                     timepoint1))) %>% 
   mutate(timepoint2 = ifelse(timepoint2 %in% c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa"), "sCST1Sa_H_sCST1Sa_M_sCST2Sa", 
                              ifelse(timepoint2 %in% c("sCST2Sc_H", "sCST2Sc", "sCST2Sh"), "sCST2Sc_H_sCST2Sc_sCST2Sh",
                                     timepoint2)))  %>% 
  mutate(timepoint3 = ifelse(timepoint3 %in% c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa"), "sCST1Sa_H_sCST1Sa_M_sCST2Sa", 
                              ifelse(timepoint3 %in% c("sCST2Sc_H", "sCST2Sc", "sCST2Sh"), "sCST2Sc_H_sCST2Sc_sCST2Sh",
                                     timepoint3))) 
  
temp_timepoint5_D <- as.matrix(temp_timepoint5_D)

chain_D = markovchainFit(temp_timepoint5_D) # This calculates the average of tp1 to tp2 and tp2 to tp3 transitions, which is what we want

markov_Cons = data.frame(chain_D$estimate@transitionMatrix)
markov_Cons_round = round(markov_Cons, digits = 2)
markov_Cons_round <- markov_Cons_round %>% 
  dplyr::rename("sCST1Sa_H\nsCST1Sa_M\nsCST2Sa" = "sCST1Sa_H_sCST1Sa_M_sCST2Sa", "sCST2Sc_H\nsCST2Sc\nsCST2Sh" = "sCST2Sc_H_sCST2Sc_sCST2Sh")

#Table of transition probabilities
markov_Cons_round %>% kable() %>% kable_classic()

#Plot
plotmat(as.matrix(t(markov_Cons_round)), pos = c(2,2), lwd = 1.5, self.lwd = 1.5, box.lwd = 1, cex.txt = 0.7, box.size = 0.1, box.type = "circle", box.prop = 1, box.col = c("#f2ad00","#ac8cae","#82a1c1", "#a7bf90"), arr.length = 0.15, arr.width = 0.15, self.cex = 0.54, self.shifty = c(0,0,0,0), self.shiftx = c(-0.14,0.14,-0.14, 0.14), shadow.size=0, box.cex = 0.9, arr.type = "triangle")
```



## Time between sampling time points  
Here, we calculate the average number of days between the sampling time points.  

```{r}
count_metadata_sample_time <- count_metadata %>% filter(population == "Denmark") %>% 
  mutate(sample_collection_date = as.Date(sample_collection_date,format = "%d-%b-%y")) %>% 
  select(Participant_ID, sample_collection_date, timepoint) %>% 
  pivot_wider(id_cols = Participant_ID, names_from = timepoint, values_from = sample_collection_date) %>% 
  mutate(time1 = as.numeric(difftime(timepoint2, timepoint1, units = c("days"))), 
         time2 = as.numeric(difftime(timepoint3, timepoint2, units = c("days")))) %>% 
  filter(!is.na(time1))

dim(count_metadata_sample_time) #should be 217 unique IDs with at least tp1 and tp2
```

Format data for plotting and calculate average distance between sampling time points in days.  
```{r}
count_metadata_sample_time_long <- count_metadata_sample_time %>%
  select(Participant_ID, time1, time2) %>% 
  pivot_longer(cols = c(time1, time2), names_to = "time_dist", values_to = "days_difference") %>% 
  filter(!is.na(days_difference))

count_metadata_sample_time_long %>% group_by(time_dist) %>% dplyr::summarise(mean = mean(days_difference), median = median(days_difference), min = min(days_difference), max = max(days_difference)) %>% kable() %>% kable_classic()
```


Plot  
```{r}
average_plot = ggplot(data = count_metadata_sample_time_long, aes(x = time_dist, y = days_difference)) +
  geom_boxplot(aes(fill = time_dist), alpha = 1, outlier.alpha = 0, size=1.5) +
  geom_point(position = position_dodge(width = 0.3), aes(color = time_dist, group = Participant_ID), alpha = 0.5) +
  #stat_boxplot(geom ="errorbar") +
  geom_line(aes(group = Participant_ID), alpha = 0.25, position = position_dodge(0.3)) +
  ylab("Days between sampling") +
  scale_x_discrete(name = "Time point",labels = c("time1" = "1 vs 2", "time2" = "2 vs 3")) +
  theme_cowplot() +
  scale_color_manual(values= c("darkgrey", "black")) +
  scale_fill_manual(values= c("white", "darkgrey")) +
  theme(legend.position = "")
average_plot
```



# Analysis of absolute abundance


### Import further metadata for the cross-sectional samples including:   
- absolute bacterial abundance ("per_swab")  
- relative staphylococcal abundance based on 16S sequencing ("staph_percentage")  
- absolute staphylococcal abundance ("staph_absolute")  
- absolute abundance of bacteria other than staphylococci ("non_staph_absolute")  
```{r}
count_metadata2 <- read.csv("data_input/count_metadata2.csv", row.names = 1)

count_metadata2 <- count_metadata2 %>% rownames_to_column("Sample") %>% left_join(count_metadata %>% filter(data_type=="cross sectional") %>% rownames_to_column("Sample")) %>% column_to_rownames("Sample")

levels(count_metadata2$sCST) = c("sCST1Sa_H", "sCST1Sa_M", "sCST2Sa", "sCST2Sc_H", "sCST2Sl", "sCST2Sh","sCST2Sc", "sCST2Se")
```


Subset samples with absolute abundance data available and bacterial CST info available (846 samples)
```{r}
count_metadata2_abs <- count_metadata2 %>% filter(!is.na(per_swab)) %>% filter(!is.na(bacterial_CST))

dim(count_metadata2_abs)
```


## Median staphylococcal relative abundance in the nasal microbiota per sCST  
```{r}
staph_median = count_metadata2_abs  %>% group_by(sCST) %>% dplyr::summarise(percentage_median = median(staph_percentage))

staph_median %>% kable() %>% kable_classic()
```


```{r fig.height=5, fig.width=10}
#Add sCST count
CST_count <- count_metadata2_abs  %>% group_by(sCST) %>% dplyr::summarise(count=n())

staph_percentage_plot = ggplot(count_metadata2_abs, aes(x = sCST, y = staph_percentage)) +
  geom_boxplot(aes(fill = sCST), alpha = 1, outlier.alpha = 0) +
  geom_point(position = position_jitterdodge(seed = 1, jitter.width = 1), aes(fill = sCST), color = "black", pch = 21,alpha = 0.6) +
  geom_text(data = CST_count, aes(y = 102, label = paste0("n=",count)), size = 4, vjust = -0.7) +
  scale_color_manual(values= sCST_colors) + 
  scale_fill_manual(values= sCST_colors) + 
  theme_cowplot() +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 9)) +
  xlab("sCST") +
  ylab("Staphylococcal relative abundance,%") +
  theme(legend.position = "", axis.title.y = element_text(size = 12)) +
  background_grid() +
  ylim(0,105)
 staph_percentage_plot
```

### Statistics on staphylococcal percentage  
We see an overall significant difference between the sCSTs regarding staphylococcal percentage  
```{r}
kruskal.test(staph_percentage ~ sCST, data = count_metadata2_abs)
```

Pairwise comparison between the sCSTs   
```{r}
pairwise.wilcox.test(count_metadata2_abs$staph_percentage, count_metadata2_abs$sCST, p.adjust.method = "BH")
```


## Absolute bacterial abundance, staphylococcal abundance, *S. aureus* abundance and *S. epidermidis* abundance
```{r}
count_table_t <- t(count_table) %>% as.data.frame()

count_metadata2_abs2 <- count_metadata2_abs %>% rownames_to_column("Sample") %>% 
  left_join(count_table_t %>%  rownames_to_column("Sample") %>% select(Sample, Staphylococcus_aureus, Staphylococcus_epidermidis), by="Sample") %>% column_to_rownames("Sample") %>% 
  mutate(aureus_absolute = (staph_absolute * Staphylococcus_aureus) /100) %>% 
  mutate(epidermidis_absolute = (staph_absolute * Staphylococcus_epidermidis) /100)
```

### Obtain abundance medians for each sCST  
```{r}
count_metadata2_abs2_median = count_metadata2_abs2 %>% group_by(sCST) %>% 
  dplyr::summarise(
    "Absolute bacterial \nabundance (*10^5)" = paste(round(median(per_swab/1e+05), digits = 2), "(IQR:",round(quantile(per_swab/1e+05,probs = 0.25),1),"-", round(quantile(per_swab/1e+05,probs = 0.75),1),")"), 
    "Absolute staphylococcal \nabundance (*10^5)" = paste(round(median(staph_absolute/1e+05), digits = 2), "(IQR:",round(quantile(staph_absolute/1e+05,probs = 0.25),1),"-", round(quantile(staph_absolute/1e+05,probs = 0.75),1),")"), 
    "Absolute S. aureus \nabundance" = paste(round(median(aureus_absolute), digits = 0), "(IQR:",round(quantile(aureus_absolute,probs = 0.25),0),"-", round(quantile(aureus_absolute,probs = 0.75),0),")"), 
    "Absolute S. epidermidis \nabundance (*10^4)" = paste(round(median(epidermidis_absolute/1e+04), digits = 2), "(IQR:",round(quantile(epidermidis_absolute/1e+04,probs = 0.25),1),"-", round(quantile(epidermidis_absolute/1e+04,probs = 0.75),1),")"))

count_metadata2_abs2_median %>% kable() %>% kable_classic()
```

### Plot absolute abundances (qPCR data)

```{r fig.height=5, fig.width=12}
count_metadata2_abs2_melt = melt(count_metadata2_abs2 %>% select(sCST, per_swab, staph_absolute, aureus_absolute, epidermidis_absolute))
count_metadata2_abs2_melt = count_metadata2_abs2_melt[which(count_metadata2_abs2_melt$value>0),]

count_metadata2_abs2_melt_medians <- count_metadata2_abs2_melt %>% group_by(sCST, variable) %>% dplyr::summarise(median=median(value))
```


```{r fig.height=5, fig.width=12}
qPCR_plot3 = ggplot() +
  
  geom_bar(data=count_metadata2_abs2_melt_medians %>% filter(variable %in% c("per_swab")), aes(x=sCST, y=median, fill=variable), stat="identity", width=0.70, position = position_nudge(x = -0.15)) +
  
   geom_bar(data=count_metadata2_abs2_melt_medians %>% filter(variable %in% c("staph_absolute")), aes(x=sCST, y=median, fill=variable), stat="identity", width=0.65, position = position_nudge(x = -0.075)) +

   geom_bar(data=count_metadata2_abs2_melt_medians %>% filter(variable %in% c("aureus_absolute", "epidermidis_absolute")), aes(x=sCST, y=median, fill=variable), stat="identity", width=0.6) +
  
scale_y_continuous("median 16S rRNA gene copies per swab") +
  
 scale_fill_manual(name = "qPCR",
                    label = c(bquote(atop("Bacterial", " absolute abundance")),
                              bquote(atop("Staphylococcal", " absolute abundance")),
                              bquote(atop(italic("S. aureus")," absolute abundance"),),
                              bquote(atop(italic("S. epidermidis"), " absolute abundance"))),
                    values = c("grey","#595959","#f2ad00","#82a1c1")) +
  

  theme_cowplot()+

  theme(legend.position = "right", legend.key.size = unit(1.2, "cm"),legend.text.align = 0,axis.title.x = element_blank()) +
 ylab("median 16S rRNA gene copies per swab") +
  background_grid()
qPCR_plot3
```

### Statistics on absolute bacterial abundance  
We see a significant difference between the sCSTs regarding absolute bacterial abundance.  
```{r}
kruskal.test(per_swab ~ sCST, data = count_metadata2_abs2)
pairwise.wilcox.test(count_metadata2_abs2$per_swab, count_metadata2_abs2$sCST, p.adjust.method = "BH")
```

### Statistics on staphylococcal absolute abundance  
There was a significant difference between the sCSTs regarding staphylococcal absolute abundance.  
```{r}
kruskal.test(staph_absolute ~ sCST, data = count_metadata2_abs2)
pairwise.wilcox.test(count_metadata2_abs2$staph_absolute, count_metadata2_abs2$sCST, p.adjust.method = "BH")
```

### Statistics on *S. aureus* absolute abundance  
There was a significant difference between the sCSTs regarding *S. aureus* absolute abundance.  
```{r}
kruskal.test(aureus_absolute ~ sCST, data = count_metadata2_abs2)
pairwise.wilcox.test(count_metadata2_abs2$aureus_absolute, count_metadata2_abs2$sCST, p.adjust.method = "BH")
```

### Statistics on *S. epidermidis* absolute abundance
There was a significant difference between the sCSTs regarding *S. epidermidis* absolute abundance.  
```{r}
kruskal.test(epidermidis_absolute ~ sCST, data = count_metadata2_abs2)
pairwise.wilcox.test(count_metadata2_abs2$epidermidis_absolute, count_metadata2_abs2$sCST, p.adjust.method = "BH")
```

#### Combined plots of staphylococcal percentage and absolute abundances (qPCR)  
```{r fig.height=10, fig.width=15}
qPCR_grid = plot_grid(qPCR_plot3,staph_percentage_plot,ncol = 1, align = "v",rel_heights = c(5,4), labels = "AUTO",label_size = 16)
qPCR_grid
```



## Carrier status based on S. aureus absolute abundance in DK (similar to Figure S3)
```{r}
df_aureus_abs <- count_metadata2_abs2 %>%
  select(Participant_ID, timepoint, data_type, aureus_absolute, carrier)
```

Fewer samples since not all have qPCR data.  
```{r}
prev_table_abs <- df_aureus_abs %>%  filter(data_type == "cross sectional") %>% 
  dplyr::summarise(S.aureus_absent_n = matrixStats::count(aureus_absolute == 0) , S.aureus_present_n= matrixStats::count(aureus_absolute>0)) %>% 
  dplyr::summarise("S.aureus_present >0" = paste0(S.aureus_present_n, " (", round(S.aureus_present_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)"),
            "S.aureus_absent (0)" = paste0(S.aureus_absent_n, " (", round(S.aureus_absent_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)")) 

prev_table_abs %>%
  kable() %>% kable_classic()
```





Which S. aureus absolute abundance would be required to detect 43% of samples as positive? (correponding to culture-positive prevalence) 
```{r}
round2 = function(x, n) {
  posneg = sign(x)
  z = abs(x)*10^n
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^n
  z*posneg
}

# Calculate exact abs abund for 43%

n <- round2((length(unique(df_aureus_abs$Participant_ID))/100)*43,0) #364 persons correspond to 43% of the cohort
df_aureus_abs_sorted_decr <- df_aureus_abs %>% arrange(-aureus_absolute) %>% rownames_to_column("Sample")
abd_43 <- df_aureus_abs_sorted_decr[n,"aureus_absolute"]
abd_43

prev_table_abs_c <- df_aureus_abs %>%  filter(data_type == "cross sectional") %>% 
  dplyr::summarise(S.aureus_absent_n = matrixStats::count(aureus_absolute <abd_43) , S.aureus_present_n= matrixStats::count(aureus_absolute >=abd_43)) %>% 
  dplyr::summarise("S.aureus_present >=1,671" = paste0(S.aureus_present_n, " (", round(S.aureus_present_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)"),
            "S.aureus < 1,671" = paste0(S.aureus_absent_n, " (", round(S.aureus_absent_n/sum(S.aureus_absent_n, S.aureus_present_n)*100),"%)")) 

prev_table_abs_c %>%
  kable() %>% kable_classic()
```


```{r fig.height=6, fig.width=8}
df_aureus_abs_sorted <- df_aureus_abs %>% arrange(aureus_absolute) %>% rownames_to_column("Sample") 
```

```{r fig.height=5, fig.width=10}
#need to take out everything below 1, otherwise negative log10
abs_cul <- ggplot(df_aureus_abs_sorted %>% filter(aureus_absolute>1), aes(x=Sample, y=aureus_absolute, fill=carrier)) +
geom_bar(position = "stack", stat = "identity", width = 1) +
scale_fill_manual(values=c("darkgrey", "#1f78b4")) +
scale_x_discrete(limits=df_aureus_abs_sorted$Sample) +  #This way still all samples included
theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none") + 
scale_y_continuous(expand = c(0,0.5), trans = "log10", name = "16S rRNA gene copies per swab", breaks = c(10,1000,100000,10000000,1000000000),labels = trans_format("log10", math_format(10^.x))) +
  theme(legend.position = "bottom", legend.key.size = unit(1.2, "cm"),legend.text.align = 0) +
  annotation_logticks(sides = "l") +
  background_grid() + 
#geom_hline(yintercept = 10, color="red", size=0.5) + 
geom_vline(xintercept = df_aureus_abs_sorted$Sample[df_aureus_abs_sorted$aureus_absolute >0][1], size=0.5) + 
geom_vline(xintercept = df_aureus_abs_sorted$Sample[df_aureus_abs_sorted$aureus_absolute >=1700][1], size=0.5) + 
annotate(x= df_aureus_abs_sorted$Sample[df_aureus_abs_sorted$aureus_absolute >0][280], y= 10, label=paste0("S. aureus present >0 16S rRNA gene copies per swab:", "n=", prev_table_abs$`S.aureus_present >0`), "label") +
annotate(x= df_aureus_abs_sorted$Sample[df_aureus_abs_sorted$aureus_absolute >=abd_43][200], y= 900, label=paste0("S. aureus present >=1,671 \n16S rRNA gene copies per swab: \n", "n=", prev_table_abs_c$`S.aureus_present >=1,671`), "label")  +
labs (x= "Study participants", title = "Estimated absolute abundance of S. aureus", fill="carrier acc. to culture")

abs_cul
```


How well are carrier status and S. aureus copy number correlated? 

Testing culture positive vs negative against each other: 
```{r}
wil <- pairwise.wilcox.test(df_aureus_abs_sorted$aureus_absolute, df_aureus_abs_sorted$carrier,paired=FALSE, p.adjust.method = "none")
wil
```
Calculate effect size
```{r}
wilcoxonR(df_aureus_abs_sorted$aureus_absolute, df_aureus_abs_sorted$carrier)
```

Median abs S. aureus abundance in Carriers vs non-carriers
```{r}
count_metadata2_abs2_median = count_metadata2_abs2 %>% group_by(carrier) %>% 
  dplyr::summarise(
    "Absolute S. aureus \nabundance" = paste(round(median(aureus_absolute), digits = 0), "(IQR:",round(quantile(aureus_absolute,probs = 0.25),0),"-", round(quantile(aureus_absolute,probs = 0.75),0),")"))

count_metadata2_abs2_median %>% kable() %>% kable_classic()
```
  
  

# Analysis of association between bacterial CSTs and sCSTs  
```{r}
count_na_removed_CS <- count_metadata2_abs %>% mutate(bacterial_CST = as.factor(bacterial_CST))


#rename:
count_na_removed_CS <- count_na_removed_CS %>% 
            mutate(bacterial_CST = case_when(
  bacterial_CST =="1a S. aureus-high" ~ "CST1a",
  bacterial_CST =="1b S. aureus-med" ~"CST1b",
  bacterial_CST =="2 Entero" ~"CST2",
  bacterial_CST =="3 S. epi" ~"CST3",
  bacterial_CST =="4 Cutibacterium"  ~"CST4",
  bacterial_CST =="5a C. accolens" ~"CST5a",
  bacterial_CST =="5b C. segment"  ~"CST5b",
  bacterial_CST =="6a M. cat/nonliq" ~"CST6a",
  bacterial_CST =="6b M. nonliq"  ~"CST6b", 
  bacterial_CST =="7 Dp/C. pseudo"  ~"CST7", 
  bacterial_CST =="8 Haemophilus" ~"CST8",
  bacterial_CST =="9a Finegoldia/Pep/Anaero" ~"CST9a",
  bacterial_CST =="9b Bacillus"~"CST9b"
            ))


dim(count_na_removed_CS)

count_na_tally = count_na_removed_CS %>% 
  group_by(sCST, bacterial_CST) %>% tally() %>% 
  mutate(sCST_sum=sum(n)) %>%
  group_by(bacterial_CST) %>% 
  mutate("percentage (sCST/CST)" = round(n/sum(n)*100,1), cst_sum = sum(n)) %>% 
  arrange(bacterial_CST, desc("percentage (sCST/CST)")) %>% 
  group_by(sCST) %>% 
  mutate("percentage (CST/sCST)" = round(n/sum(n)*100, 1))

count_na_tally %>% kable() %>% kable_classic()
```


### Bacterial CST vs sCST statistics
This shows that there is a significant difference between sCSTs with regards to which bacterial CSTs they are associated with.  
```{r}
chisq.test(table(count_na_removed_CS$sCST, count_na_removed_CS$bacterial_CST))
```



## Bacterial CSTs of excluded samples
This shows the bacterial CST assignments of the samples that have been excluded due to insufficient read count. This is to assess whether the low *tuf* gene read count might be due to technical reasons or an expected low staphylococcal abundance in the samples.  
CAVE: None of these samples are grouped into bacterial CST2 (Enterobacteriaceae) and CST8 (Haemophilus) -> Add to table manually
```{r fig.height=3, fig.width=4.5}
# subset to those that have a bacterial CST assignment (n=128)
count_temp_removed = count_meta_removed %>% filter(!is.na(bacterial_CST))

count_temp_removed <- count_temp_removed %>% 
            mutate(bacterial_CST = case_when(
  bacterial_CST =="1a S. aureus-high" ~ "CST1a",
  bacterial_CST =="1b S. aureus-med" ~"CST1b",
  bacterial_CST =="2 Entero" ~"CST2",
  bacterial_CST =="3 S. epi" ~"CST3",
  bacterial_CST =="4 Cutibacterium"  ~"CST4",
  bacterial_CST =="5a C. accolens" ~"CST5a",
  bacterial_CST =="5b C. segment"  ~"CST5b",
  bacterial_CST =="6a M. cat/nonliq" ~"CST6a",
  bacterial_CST =="6b M. nonliq"  ~"CST6b", 
  bacterial_CST =="7 Dp/C. pseudo"  ~"CST7", 
  bacterial_CST =="8 Haemophilus" ~"CST8",
  bacterial_CST =="9a Finegoldia/Pep/Anaero" ~"CST9a",
  bacterial_CST =="9b Bacillus"~"CST9b"
            ))


removed_summary = count_temp_removed %>% group_by(bacterial_CST) %>% 
  tally() %>%
  add_row(bacterial_CST = "CST2", n=0) %>% #add CST2 manually (0 occurrences)
  add_row(bacterial_CST = "CST8", n=0) %>% #add CST8 manually (0 occurrences)
  mutate("percentage \n(excluded samples/CST)" = round(n/sum(n)*100,1), sum = sum(n))

removed_summary %>% kable() %>% kable_classic()
```


## Mosaic plot
```{r fig.height=10, fig.width=10}
# Create a mosaic plot using vcd
# On pooled sCST data with all CSTs
count_na_removed_CS_pooled <- count_na_removed_CS %>% 
  mutate(sCST_pooled =case_when(
     sCST %in% c("sCST1Sa_H","sCST1Sa_M") ~"1.sCST1",
     sCST == "sCST2Sa" ~"2.sCST2Sa",
     sCST == "sCST2Se" ~"3.sCST2Se",
    .default = "4.sCST2 (other)"))

mosaic(~ bacterial_CST + sCST_pooled,data=count_na_removed_CS_pooled, 
       shade = TRUE, 
       legend = TRUE,
       rot_labels=c(45,0),
       offset_label =c(0,0),
       zero_size=0,
       direction="v",
       labeling=labeling_residuals,
       gp=shading_hcl)
```

```{r}
table(count_na_removed_CS$sCST)
table(count_na_removed_CS$bacterial_CST)
table(count_na_removed_CS$sCST, count_na_removed_CS$bacterial_CST)
```


Keep any CSTs with any correlation < -2.5 or >2.5 and n=>30 and pool CoNS sCSTs

```{r fig.height=10, fig.width=10}
mosaic(~ bacterial_CST + sCST_pooled,data=count_na_removed_CS_pooled, 
       shade = TRUE, 
       legend = TRUE,
       rot_labels=c(0,0),
       offset_label =c(0,0),
       zero_size=0,
       direction="v",
       labeling=labeling_residuals,
       subset=bacterial_CST %in% c("CST1a", "CST3", "CST4", "CST5a", "CST7", "CST9a"),
       gp=shading_hcl)

```



#### Plot bacterial CST assignments of the samples that have been excluded due to insufficient read count.  
```{r fig.height=4, fig.width=12}
removed_CST = ggplot(removed_summary, aes(x = bacterial_CST, y = `percentage 
(excluded samples/CST)`, fill = bacterial_CST)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values= CST_colors, labels = c(
    "CST1a - S. aureus (high density)",
    "CST1b - S. aureus (medium density)",
    "CST2  -  Enterobacteriaceae",
    "CST3  -  S. epidermidis",
    "CST4  -  Cutibacterium",
    "CST5a - Corynebacterium accolens",
    "CST5b - Corynebacterium segmentosum",
    "CST6a - Moraxella catarrhalis/\n              Moraxella nonliquefaciens",
    "CST6b - Moraxella nonliquefaciens",
    "CST7  -  Dolosigranulum pigrum/\n              Corynebacterium pseudodiphteriticum",
    "CST8  -  Haemophilus",
    "CST9a - Finegoldia/Peptoniphilus/\n              Anaerococcus",
    "CST9b - Bacillus"),
    name = "CST") +
  scale_y_continuous(expand = c(0,0)) +
  theme_cowplot() +
  background_grid(major = "y") +
  ylab("Percentage of samples [%]") + 
  xlab("") 
removed_CST
```


Determine the number of true staphylococci negatives - i.e. which of the 169 samples with insufficient *tuf* read count also have no *Staphylococcus* genus reads in the 16S data?   
```{r}
count_meta_removed %>% filter(!is.na(staph_percentage)) %>% dplyr::count() %>% dplyr::rename(Number_of_samples_with_16S_data = n) 

count_meta_removed %>% filter(staph_percentage == 0) %>% dplyr::count() %>%  dplyr::rename(Number_of_Staph_negatives = n) 
```



# Demographic factors    

## Population 

## Sex  


## Distribution of sexes in the two major groups (sCST1 and sCST2)  
#### Test whether the two major sCST groups differ with regards to distribution of sexes     
```{r}
meta_sex <- count_metadata2 %>% filter(!is.na(sex)) %>%  mutate(sCST_main = ifelse(sCST %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", "sCST2")) 

chisq.test(table(meta_sex$sex,meta_sex$sCST_main))
```

#### Calculate percentages and 95% CI of sex distribution within the two major sCST groups  
```{r}
meta_sex <- meta_sex %>% group_by(sex, sCST_main) %>% tally() %>% group_by(sex) %>% mutate(total_per_sex = sum(n), perc = round((n/total_per_sex)*100,1))

meta_sex$row.no<-1:nrow(meta_sex)
df_ci<-data.frame()
n<-1:nrow(meta_sex)
for(i in n){
  ci_low<-round2(binom.test(meta_sex$n[i],meta_sex$total_per_sex[i])$conf.int[1]*100,1)
  ci_hi<-round2(binom.test(meta_sex$n[i],meta_sex$total_per_sex[i])$conf.int[2]*100,1)
  row.no<-i
  ci_com<-data.frame(row.no,ci_low,ci_hi)
  df_ci<-rbind(df_ci,ci_com)
}

meta_sex %>% 
  left_join(df_ci,by="row.no") %>%
  select(-row.no) %>% kable() %>% kable_classic()
```


#### Calculate percentage of men and women in each sCST normalized for an overall differing number of samples in each sex
```{r eval=FALSE, fig.height=5, fig.width=9, warning=FALSE}
metadata_tally = na.exclude(count_metadata2 %>% group_by(sCST, sex) %>% tally())
metadata_tally = metadata_tally %>% group_by(sex) %>% mutate(normalise_n = n/sum(n)) %>% group_by(sCST) %>% mutate(percentage = round(normalise_n/sum(normalise_n)*100, digit = 1))

metadata_tally %>% kable() %>% kable_classic()
```

#### Test whether the sCSTs differ with regards to distribution of sexes  
```{r eval=FALSE, fig.height=5, fig.width=9, warning=FALSE}
chisq.test(table(count_metadata2$sCST,count_metadata2$sex))
```

## Distibution of sexes in the Danish and the German population  
Test whether the populations differ with regards to distribution of sexes  
```{r}
table(count_metadata2$population,count_metadata2$sex)
chisq.test(table(count_metadata2$population,count_metadata2$sex))
```


## Distibution of sexes in the sCSTs between the Danish and the German population  
Calculate percentage of men and women in each sCST in the two populations normalized for a differing number of samples in the two populations and an overall different number of men vs women  
```{r fig.height=7, fig.width=12, warning=FALSE}
count_metadata2$sex <- factor(count_metadata2$sex, levels= c("Male", "Female"))

metadata_tally = na.exclude(count_metadata2 %>% group_by(sCST, sex) %>% tally())
metadata_tally = metadata_tally %>% group_by(sex) %>% mutate(normalise_n = n/sum(n)) %>% group_by(sCST) %>% mutate(percentage = round(normalise_n/sum(normalise_n)*100, digit =1))
metadata_tally %>% kable() %>% kable_classic()
```


### Plot distribution of sexes in the sCSTs in the two populations (pooled)
```{r fig.height=4, fig.width=12, warning=FALSE}
sex_plot = ggplot(metadata_tally, aes(x = sCST, y = normalise_n, fill = sex)) +
  geom_bar(position = "fill", stat = "identity") +
 
  geom_hline(yintercept = 0.5, linetype ="dashed") +
  scale_fill_manual(name = "Sex", values = c(Male = "#446454", Female = "#a89985")) + 
  scale_y_continuous(name = "Percentage of samples",expand = c(0,0), labels = scales::label_percent()) +
  theme_cowplot() +
  theme(axis.title.x = element_blank())
sex_plot
```


## Age 


### Plot number of samples of each age group and sex within the two populations
```{r fig.height=7, fig.width=9, warning=FALSE}
metadata_tally = count_metadata2 %>% group_by(population, age, sex) %>% tally()
metadata_tally = na.exclude(metadata_tally)

population_dist = ggplot(metadata_tally, aes(x = age, y = n, fill = sex)) +
  geom_bar(position = "stack", stat = "identity") +
  facet_grid(vars(sex),vars(population)) +
  scale_fill_manual(name = "Sex", values = c(Male = "#446454", Female = "#a89985")) +
  scale_y_continuous(expand = c(0,5)) +
  ylab("Number of samples") +
  xlab("Age group") +
  coord_flip() +
  theme_cowplot() +
  panel_border() + background_grid()
population_dist
```


### Calculate median age group per cohort and test whether the Danish and German cohorts are different with regards to participants' age.  
```{r}
count_metadata_test_age <- count_metadata2 %>% mutate(age_group=case_when(
     age ==  "< 25" ~1,
     age =="25-29"~2,
     age =="30-34"~3,
     age =="35-39"~4,
     age =="40-44"~5,
     age =="45-49"~6,
     age =="50-54"~7,
     age =="55-59"~8,
     age ==" 60"~9))

count_metadata_test_age %>% group_by(population) %>% dplyr::summarise(median(age_group, na.rm = T), min(age_group, na.rm = T), max(age_group, na.rm = T)) %>% kable() %>% kable_classic()

chisq.test(table(count_metadata2$population,count_metadata2$age))
```

### Visualize age distribution by sCST normalized for a different number of samples in each age group  
```{r fig.height=5, fig.width=12}
count_metadata2$age <- factor(count_metadata2$age, levels=c("< 25", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59"," 60"))

metadata_tally = na.exclude(count_metadata2 %>% group_by(sCST, age) %>% tally())
metadata_tally = metadata_tally %>% group_by(age) %>% mutate(normalise_n = n/sum(n)) %>% group_by(sCST) %>% mutate(percentage = round(normalise_n/sum(normalise_n)*100, digit = 1))
metadata_tally %>% kable() %>% kable_classic()

age_plot_alt2 = ggplot(metadata_tally, aes(x = sCST, y = normalise_n, fill = age)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(name = "Percentage of samples",expand = c(0,0), labels = scales::label_percent()) +
  scale_x_discrete(name = "Age group",labels = function(x) str_wrap(x, width = 9)) +
  scale_fill_viridis(name= "Age group", option="cividis", discrete=TRUE, direction = -1) +
  theme_cowplot() 

age_plot_alt2
```

## Smoking  


### Distribution of smoking status in the two major groups (sCST1 and sCST2)  
#### Test whether the two major sCST groups differ with regards to distribution of smoking status    
```{r fig.width=5, warning=FALSE}
count_metadata2_smo <- count_metadata2 %>% mutate(sCST_main = ifelse(sCST %in% c("sCST1Sa_H", "sCST1Sa_M"), "sCST1", "sCST2"))

chisq.test(table(count_metadata2_smo$sCST_main, count_metadata2_smo$smoker))
```

#### Calculate percentages and 95% CI of smoking status distribution within the two major sCST groups  
```{r fig.width=5, warning=FALSE}
count_metadata2_smo <- count_metadata2_smo %>% filter(!is.na(smoker)) %>% group_by(smoker, sCST_main) %>% tally() %>% group_by(smoker) %>% mutate(total_per_population = sum(n), perc = round((n/total_per_population)*100,1))


count_metadata2_smo$row.no<-1:nrow(count_metadata2_smo)
df_ci<-data.frame()
n<-1:nrow(count_metadata2_smo)
for(i in n){
  ci_low<-round2(binom.test(count_metadata2_smo$n[i],count_metadata2_smo$total_per_population[i])$conf.int[1]*100,1)
  ci_hi<-round2(binom.test(count_metadata2_smo$n[i],count_metadata2_smo$total_per_population[i])$conf.int[2]*100,1)
  row.no<-i
  ci_com<-data.frame(row.no,ci_low,ci_hi)
  df_ci<-rbind(df_ci,ci_com)
}

count_metadata2_smo %>% 
  left_join(df_ci,by="row.no") %>%
  select(-row.no) %>% kable() %>% kable_classic()
```


#### Test whether the sCSTs differ regarding smoking status of participants  
```{r fig.width=5, warning=FALSE}
count_metadata2$smoker <- factor(count_metadata2$smoker,levels=c("Yes", "No"))

chisq.test(table(count_metadata2$sCST, count_metadata2$smoker))
```

### Visualize distribution of smoking status by sCST (normalized for overall different number of smokers vs non-smokers)  
```{r fig.height=5, fig.width=12, warning=FALSE}
metadata_tally = na.exclude(count_metadata2 %>% group_by(sCST, smoker) %>% tally())
metadata_tally = metadata_tally %>% group_by(smoker) %>% mutate(normalise_n = n/sum(n)) %>% group_by(sCST) %>% mutate(percentage = round(normalise_n/sum(normalise_n)*100, digit = 1))
metadata_tally %>% kable() %>% kable_classic()

smoker_plot = ggplot(metadata_tally, aes(x = sCST, y = normalise_n, fill = smoker)) +
  geom_bar(position = "fill", stat = "identity") +
  geom_hline(yintercept = c(0.5), linetype ="dashed") +
  scale_fill_manual(name = "Smoker", values = c(No = "#a89985",Yes = "#446454")) + 
  scale_y_continuous(expand = c(0,0), labels = scales::label_percent()) +
  xlab("sCST") +
  ylab("Percentage of samples") +
  theme_cowplot()
smoker_plot
```

### Distribution of smoking status between the populations  
```{r}
table(count_metadata2$population,count_metadata2$smoker)
chisq.test(table(count_metadata2$population,count_metadata2$smoker))
```




# Culturing results for S. aureus (Danish cross-sectional samples)  
#### Test whether sCSTs differ according to S. aureus culturing status  
```{r warning=FALSE}
count_metadata2_DK = count_metadata2 %>% filter(!is.na(carrier)) #excludes the German samples (no culture data)

count_metadata2_DK <- count_metadata2_DK %>% mutate(carrier=case_when(carrier=="Yes" ~ "Positive", carrier=="No" ~ "Negative"))

count_metadata2_DK$carrier <- factor(count_metadata2_DK$carrier, levels = c("Positive", "Negative"))

metadata_tally = count_metadata2_DK %>% group_by(sCST, carrier) %>% tally()

metadata_tally %>% group_by(sCST) %>% mutate(percentage = round(n/sum(n)*100,digits = 1), group_sum = sum(n)) %>% kable() %>% kable_classic()

chisq.test(table(count_metadata2_DK$sCST,count_metadata2_DK$carrier))
```
How many carriers overall 
```{r}
table(count_metadata2_DK$carrier)
```



```{r fig.height=5, fig.width=12, warning=FALSE}
carrier_plot = ggplot(metadata_tally, aes(x = sCST, y = n, fill = carrier)) +
  geom_bar(position = "fill", stat = "identity") +
  geom_hline(yintercept = c(0.25,0.75), linetype ="dashed") +
  scale_fill_manual(name = "S. aureus \nculture status", values = c(Negative = "#a89985", Positive = "#446454")) + 
  scale_y_continuous(name = "Percentage", expand = c(0,0), labels = scales::label_percent()) +
  theme_cowplot() +
  theme(axis.title.x = element_blank())
  carrier_plot
```


### Semi-quantitative determination of S. aureus colonies per sample  
```{r fig.height=5, fig.width=9, warning=FALSE}
count_metadata2_DK$quantity <- factor(count_metadata2_DK$quantity, levels = c("5. >500 colonies", "4. 50-500 colonies", "3. <50 colonies", "2. Culture-positive \nafter enrichment", "1. Absence of \nany colonies"))
                                      
metadata_tally = count_metadata2_DK %>% group_by(sCST, quantity) %>% tally()
metadata_tally %>% group_by(sCST) %>% mutate(percentage = round(n/sum(n)*100, digits = 1), group_sum = sum(n)) %>% kable() %>% kable_classic()
```

Plot 
```{r fig.height=5, fig.width=12}
carrier_plot2 = ggplot(metadata_tally, aes(x = sCST, y = n, fill = quantity)) +
  geom_bar(position = "fill", stat = "identity") +
  
    scale_fill_manual(name = "S. aureus colony\nquantity per sample", values = c("#446454","#698376", "#8ea298", "#b4c1ba",  "#a89985"), labels=c(">500 colonies", "50 - 500 colonies", "<50 colonies", "Culture-positive \nafter enrichment","Absence of \nany colonies")) +
  

  scale_y_continuous(name = "Percentage", expand = c(0,0), labels = scales::label_percent()) +
  theme_cowplot() +
  guides(fill = guide_legend(byrow = TRUE)) + 
  theme(axis.title.x = element_blank())
carrier_plot2
```

